Gravitational Lens Recovery with \Glass: Measuring the mass profile and shape of a lens

### 2014-01-26 limits/ms.tex
Fixed figure 7.
Brought over Justin's changes.

\begin{abstract}
We use a new non-parametric gravitational modelling tool---\Glass---to
determine what quality of data (strong lensing, stellar kinematics, and/or
stellar masses) are required to measure the circularly averaged mass profile of
a lens and its shape. \Glass\ uses an under-constrained adaptive grid of mass
pixels to model the lens, searching through thousands of models to marginalise
over model uncertainties. Our key findings are as follows: (i) for pure lens
data, multiple sources with wide redshift separation give the strongest
constraints as this breaks the well-known mass-sheet or steepness degeneracy;
(ii) a single quad with time delays also performs well, giving a good recovery
of both the mass profile and its shape; (iii) stellar masses -- for lenses
where the stars dominate the central potential -- can also break the steepness
degeneracy, giving a recovery almost as good as having time delay data or
multiple source redshifts; (iv) stellar kinematics are similarly powerful if
the data extend beyond the half light radius of the stars $r \gg r_{1/2}$
(required to integrate out the effect of the unknown velocity anisotropy,
$\beta(r)$); and (v) if the lensing data already probe the mass profile over
the region $r \sim r_{1/2}$, then stellar kinematic data can be used to probe
$\beta(r)$ -- an interesting dynamic quantity in its own right. Where
information on the mass distribution from lensing and other probes becomes
redundant, this opens up the possibility of using strong lensing to constrain
cosmological models. We will study this, and present the first results from
\Glass\ applied to real data, in forthcoming papers.
\end{abstract}

@@

Strong gravitational lenses are rare.  The reason is that the line of
sight to a light source in the universe rarely goes through a region
such that the projected density in its sky neighbourhood exceeds the
critical density to produce multiple images.  To date, some $\sim400$
strong lenses are known,\footnote{See {\tt http://www.masterlens.org}
  for a catalogue.}  but this number is expected to increase to
several thousand over the next ten years.  The optimism derives from
new and upcoming surveys, both
ground-based\footnote{http://pan-starrs.ifa.hawaii.edu}$^{\rm,}$\footnote{http://www.darkenergysurvey.org}
and spacecraft,\footnote{http://www.euclid-ec.org} together with a
community of citizen-science volunteers examining the image data for
candidates.\footnote{http://spacewarps.org}

Since lensing depends only on gravity, strong lenses are a rather
unique window on dark matter, and indirectly on the cosmological model
as well, and hence the thousands of to-be-discovered lenses are
eagerly awaited \citep{2010CQGra..27w3001B,2012arXiv1206.1225A}.
Extracting results from lensing, however, will require modelling, and
it is towards that goal that the present work is directed.

To see why lens modelling is of crucial importance, let us recall the
essential quantities that appear in lensing.  First we have the
distances: let $D_L$, $D_S$, $D_{LS}$ be the angular-diameter
distances to the lens, source, and from lens to source; they are all
strictly $\propto c/H_0$.  Typically
\begin{equation}
D_L \approx z \frac c{H_0}\ \hbox{and}\ \frac{D_S}{D_{LS}} \sim 1.
\end{equation}
For multiple images the sky-projected density must exceed the critical
density of $D_S/D_{LS}$ times
\begin{equation}
\Sigma_L = \frac{c^2}{4\pi G D_L} \sim \frac{1\rm\,kg\,m^{-2}}{z_L}
\end{equation}
in some region.  The separation between the lensed images is of order
the Einstein radius $\theta_E$.  The image separation is related to
the mass by
\begin{equation}
\theta_E \sim \frac{R_G}{D_L} \frac{D_{LS}}{D_S}
\end{equation}
where $R_G$ is the gravitational radius $GM/c^2$.  If the source is a
quasar or otherwise rapidly variable, a time delay $\Delta t$ in the variability
will be present where
\begin{equation}
\Delta t \sim R_G/c \,.
\end{equation}
So in principle, one can not only measure the mass of the lens, one
can use the dependence on the cosmology-dependent $D$ factors to
extract the cosmological model and all its parameters.
\cite{1937ApJ....86..217Z} drew attention to the former, and
\cite{1964MNRAS.128..307R,1966MNRAS.132..101R} pointed out the latter,
all long before lenses were discovered.  The difficulty became
apparent soon after the discovery of the first lens by
\cite{1979Natur.279..381W}.  In the first ever paper on lens
modelling, \cite{1981ApJ...244..736Y} found that many plausible mass
distributions could reproduce the data.

@@

\begin{figure}
\plottwo{AAExtendedDoubleR1_TmS-shape-hist-11.pdf}{BCZContrastR1R3_TmS-shape-hist-212.pdf}
\caption{ \hilight{Larger labels.} The two plots in the left most column show sample distributions of
  $s=\lambda_1/\lambda_2$, the ratio of the largest to smallest principal axes.
  The upper (poor) plot is from the \mockAA{} galaxy, while the lower (good) plot is
  from the \mockBC{} galaxy. The black vertical lines mark the $2\sigma$
  region and the vertical blue line is the value of $s$ for the corresponding
mock galaxy. }

>>>>

\begin{abstract}
We use a new non-parametric gravitational modelling tool---\Glass---to
determine what quality of data (strong lensing, stellar kinematics, and/or
stellar masses) are required to measure the circularly averaged mass profile of
a lens and its shape. \Glass\ uses an under-constrained adaptive grid of mass
pixels to model the lens, searching through thousands of models to marginalise
over model uncertainties. Our key findings are as follows: (i) for pure lens
data, multiple sources with wide redshift separation give the strongest
constraints as this breaks the well-known mass-sheet or steepness degeneracy;
(ii) a single quad with time delays also performs well, giving a good recovery
of both the mass profile and its shape; (iii) stellar masses -- for lenses
where the stars dominate the central potential -- can also break the steepness
degeneracy, giving a recovery almost as good as having time delay data or
multiple source redshifts; (iv) stellar kinematics provide a robust measure of the mass at the half light radius of the stars $r \gg r_{1/2}$ that can also break the steepness degeneracy if $r_{1/2} \neq r_E$ -- the Einstein radius; and (v) if $r_E \sim r_{1/2}$, then stellar kinematic data can be used to probe the stellar velocity anisotropy 
$\beta(r)$ -- an interesting quantity in its own right. Where
information on the mass distribution from lensing and/or other probes becomes
redundant, this opens up the possibility of using strong lensing to constrain
cosmological models. We will study this, and present the first results from
\Glass\ applied to real data, in forthcoming papers.
\end{abstract}

@@

Strong gravitational lenses are rare. Since the discovery of the first lens Q0957+561 \citep{1979Natur.279..381W}, just $\sim400$ have been discovered to date\footnote{See e.g. {\tt http://www.masterlens.org} for a catalogue.}. However, this number is expected to increase to
several thousand over the next ten years as new surveys, both
ground-based\footnote{http://pan-starrs.ifa.hawaii.edu}$^{\rm,}$\footnote{http://www.darkenergysurvey.org}
and space-based\footnote{http://www.euclid-ec.org} -- together with a
community of citizen-science volunteers examining the image data for
candidates\footnote{{\tt http://spacewarps.org}} -- come online.

% since distant light source in the universe rarely goes through a region such that the projected density in its sky neighbourhood exceeds the critical density to produce multiple images.

Since lensing depends only on gravity, strong lenses offer a
unique window onto dark matter and cosmology \citep{2010CQGra..27w3001B,2012arXiv1206.1225A}. However, 
extracting dark matter properties or cosmological constraints from these lensing data will require sophisticated modelling. In particular, with an unprecedented data set imminent, it is prudent to look again at systematic errors in the lens models to determine what quality of data (in particular complementary data from stellar/gas kinematics, lens time delays and/or stellar mass constraints) are required to address problems of interest. It is towards that goal that this present work is directed.

To see why lens modelling details are of crucial importance, let us recall the
essential quantities that appear in lensing.  First we have the
distances. Let $D_L$, $D_S$, $D_{LS}$ be the angular-diameter
distances to the lens, source, and from lens to source; these are all proportional to $c/H_0$ (where $c$ is the speed of light in vacuo, and $H_0$ the Hubble constant), with additional terms that depend on the particular choice of cosmology \citep[e.g.][]{1999coph.book.....P}. Typically:
\begin{equation}
D_L \approx z \frac c{H_0}\ \hbox{and}\ \frac{D_S}{D_{LS}} \sim 1.
\end{equation}
where the second relation follows from the fact that the light deflection angle is maximised when the lens is equidistant between the observer and the source \citep[e.g.][]{1992grle.book.....S}.

GOT TO HERE.

For multiple images the sky-projected density must exceed the critical
density of $D_S/D_{LS}$ times
\begin{equation}
\Sigma_L = \frac{c^2}{4\pi G D_L} \sim \frac{1\rm\,kg\,m^{-2}}{z_L}
\end{equation}
in some region.  The separation between the lensed images is of order
the Einstein radius $\theta_E$.  The image separation is related to
the mass by
\begin{equation}
\theta_E \sim \frac{R_G}{D_L} \frac{D_{LS}}{D_S}
\end{equation}
where $R_G$ is the gravitational radius $GM/c^2$.  If the source is a
quasar or otherwise rapidly variable, a time delay $\Delta t$ in the variability
will be present where
\begin{equation}
\Delta t \sim R_G/c \,.
\end{equation}
So in principle, one can not only measure the mass of the lens, one
can use the dependence on the cosmology-dependent $D$ factors to
extract the cosmological model and all its parameters.
\cite{1937ApJ....86..217Z} drew attention to the former, and
\cite{1964MNRAS.128..307R,1966MNRAS.132..101R} pointed out the latter,
all long before lenses were discovered.  The difficulty became
apparent soon after the discovery of the first lens by
\cite{1979Natur.279..381W}.  In the first ever paper on lens
modelling, \cite{1981ApJ...244..736Y} found that many plausible mass
distributions could reproduce the data.

@@

\begin{figure}
\plottwo{AAExtendedDoubleR1_TmS-shape-hist.pdf}{BCZContrastR1R3_TmS-shape-hist.pdf}
\caption{ \hilight{Larger labels.} The two plots in the left most column show sample distributions of
  $s=\lambda_1/\lambda_2$, the ratio of the largest to smallest principal axes.
  The upper (poor) plot is from the \mockAA{} galaxy, while the lower (good) plot is
  from the \mockBC{} galaxy. The black vertical lines mark the $2\sigma$
  region and the vertical blue line is the value of $s$ for the corresponding
mock galaxy. }
